---
title: "Bac2 bioinformatic introduction"
author: "Dugauquier Rémy"
date: "4/19/2022"
output: html_document
---

# Introduction

Into this document you can find all Rcommand line to compute some analyzes of the internship.

# Packages installation

```{r}
install.packages("writexl")

if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("Gviz")

```

# Packages importation

```{r}
library(stringr)
library(seqinr)
library(writexl)
library(ggplot2)
library(readxl)
library(readr)
library(dplyr)
library(Gviz)
```

# Analyse of correct contigs

The goal of this part is to create a table with all our gene id sequences, with their contig id and then compute a plot of the number of genes by contigs.

So we need to create this kind of table

| gene ID | contig ID |
-----------------------
| gene A  | contig A  |
| gene B  | contig A  |
| gene C  | contig B  |

To do that we are going to use the contig ID of the correct one's that we filtered previously, and also the gene gene.

For this first part we are going to import our fasta file with all the gene id with the seqnir function and compute a table with the gene name, the length, and the sequences.

```{r}
# importation of sequence

all_gene_sequence_path <- "/Users/remydugauquier/pCloud\ Drive/thesis/intership_bac_3/bac3_internship_gene_sequences.fna"

read.fasta(all_gene_sequence_path,as.string = TRUE,forceDNAtolower = FALSE) -> all_sequence

# creation of a little for loop to extract each sequences

# this loop take sequence by sequence, extract the name and information and paste them into a table

for(i in seq(from=1, to=length(all_sequence), by=1)){
  
  # at the first round of the loop (when i = 1, we are going to create the result table)
  
  if(i == 1) {
    #We just create a dataframe to put all sequence with name
    sequence_table <- data.frame(sequence_name = "",
                                 sequence = "",
                                 length = "",
                                 annotation="",
                                 stringsAsFactors=FALSE)
    sequence_table <- sequence_table[-c(1),]
  }
  # We extract one sequence
  all_sequence[i] -> one_sequence
  # Extraction of sequence name with the function that extract the attribute (so the name) of the sequence
  attributes(one_sequence) -> one_seq_name
  one_seq_name$names -> one_seq_name
  # We put the sequence name into the table, into the first column
  one_seq_name -> sequence_table[i,1]
  # We add the annotation
  getAnnot(one_sequence) -> one_seq_annot
  one_seq_annot[[1]][1] -> sequence_table[i,4]
  # Now we need to add the sequence
  one_sequence[[1]][1]-> sequence_table[i,2]
  # And finally we want to add the length of the genes
  # we simply use the command nchar which able to count the number of character of a string, here count the number of nucleotide
  nchar(sequence_table[i,2]) -> sequence_table[i,3]
} 
```

Now we have our table almost done. We just need to add the contig id, the start position and the stop position.

So contig Id is the first part of the gene id. So we are going to extract this part from the gene id and create the variable contig id.

In R we can use the function str_split_fixed to split a name according to a letter/word/... 

```{r}

str_split_fixed(sequence_table$sequence_name, "_", 6) -> sequence_table[,c(5:10)]

paste(sequence_table$V5,sequence_table$V6,
      sequence_table$V7,sequence_table$V8,
      sequence_table$V9,sep="_") -> sequence_table$contig_id

sequence_table[,-c(5:10)] -> sequence_table
```

For start, stop and strand, we can use the same way.
But it's faster to use regular expression.

```{r}

str_extract_all(sequence_table$annotation,"(?<=start:).+(?=stop:)")-> sequence_table$start
str_extract_all(sequence_table$annotation,"(?<=stop:).+(?=length:)")-> sequence_table$stop
str_extract_all(sequence_table$annotation,"(?<=strand:).+(?=start:)")-> sequence_table$strand

# we put it as numeric

as.numeric(sequence_table$start) -> sequence_table$start
as.numeric(sequence_table$stop) -> sequence_table$stop
as.character(sequence_table$strand) -> sequence_table$strand

```


Now we have got our table, we just finish to filter it to keep only the correct contig, so the one's that we filter according to the size of contigs.

To do that we can import the contig id file, and filter the name according to our file.

```{r}
# we import the contig sequences files

all_correct_contig_path <- "/Users/remydugauquier/pCloud\ Drive/thesis/intership_bac_3/correct_contig_sequences.fasta"

read.fasta(all_correct_contig_path,as.string = TRUE,forceDNAtolower = FALSE) -> all_contig

# then we extract all the name
attributes(all_contig) -> all_contig_name

all_contig_name$names -> all_contig_name

# and we use the function subset to keep only correct contig into our table

subset(sequence_table, sequence_table$contig_id %in% all_contig_name) -> correct_sequence_table

```

So we had 670 sequences and now 634 sequences.

We can register the table for later.

```{r}

write_xlsx(correct_sequence_table,"/Users/remydugauquier/pCloud\ Drive/thesis/intership_bac_3/correct_sequence_table.xlsx")

```

To finish the analyze of correct contigs, we are going to compute a little plot to see all contigs and the number of genes that they contains.

```{r}
# we start by using table function to count the number of occurence of each contig

table(correct_sequence_table$contig_id) -> table_for_plot
as.data.frame(table_for_plot) -> table_for_plot
names(table_for_plot)[1] <- "contig_id"
names(table_for_plot)[2] <- "number_of_gene"

# now we create a barplot

ggplot(table_for_plot, aes(x = contig_id, y = number_of_gene))+
  geom_col(fill = "#0073C2FF")+
  geom_text(aes(label = number_of_gene), color = "black")+
  ggtitle("Number of genes by contigs (<1500 pb)")+
  xlab("Contigs ID")+
  ylab("Number of genes")+
  theme(axis.text.x = element_text(angle=90))

```


# Prophage prediction

Now we have our sequence table and also our prophage prediction. So we'll filter this table according to the prophage or not.

We start by import prophage table.

```{r}
library(readr)
final_viral_score <- read_delim("~/pCloud Drive/thesis/intership_bac_3/bac3_contig_prophage_result.out/final-viral-score.tsv", 
    "\t", escape_double = FALSE, trim_ws = TRUE)

library(readxl)
correct_sequence_table <- read_excel("~/pCloud Drive/thesis/intership_bac_3/correct_sequence_table.xlsx")

```


The first sequence is our contig, but the name is not correct. So we re-use split function to discard ||full.

```{r}
#
```

Then we want only the best prediction. According to our knowledge, the best score is 0.8.

So we need to subset the score.

```{r}

```

These contig are the one's detected as prophage, so we add a new column name by "prophage" into our correct_sequence_table. Then we filter our table according to this column to create an other table prophage_sequence_table that we save as excel.

To add the new column we can do a for loop again.

```{r}
# first I create a new column name prophage and I put FLASE in all row
correct_sequence_table$prophage <- FALSE

# then with a for loop, I take row by row, and if our contig id is find into the virsorter result table, I change the FALSE in TRUE
for(i in seq(from=1, to=nrow(correct_sequence_table), by=1)){
  if(correct_sequence_table[i,5] %in% unique(final_viral_score$seqname)){
    correct_sequence_table[i,9] <- TRUE
  }
}

# I filter my table according to the prophage column and I save it as excel

subset(correct_sequence_table, correct_sequence_table$prophage %in% TRUE) -> prophage_sequence_table

write_xlsx(prophage_sequence_table,"/Users/remydugauquier/pCloud\ Drive/thesis/intership_bac_3/prophage_sequence_table.xlsx")
```

To finish, we are going to re-plot this table but change the color according to prophage or not.

```{r}
table(correct_sequence_table$contig_id) -> table_for_plot
as.data.frame(table_for_plot) -> table_for_plot
names(table_for_plot)[1] <- "contig_id"
names(table_for_plot)[2] <- "number_of_gene"

table_for_plot$prophage <- FALSE

for(i in seq(from=1, to=nrow(table_for_plot), by=1)){
  if(table_for_plot[i,1] %in% unique(final_viral_score$seqname)){
    table_for_plot[i,3] <- TRUE
  }
}

ggplot(table_for_plot, aes(x = contig_id, y = number_of_gene,fill=prophage))+
  geom_bar(stat = "identity")+
  geom_text(aes(label = number_of_gene), color = "black")+
  ggtitle("Number of genes by contigs (<1500 pb)")+
  xlab("Contigs ID")+
  ylab("Number of genes")+
  theme(axis.text.x = element_text(angle=90))+scale_fill_manual(values=c( "#E69F00", "#56B4E9"))
```

What can we observe ? 

# Functionnal prediction preparation of aa sequence

For the functionnal prediction of gene with PFAM, we need to use as input our gene sequence from prophage in aa format.

So we need to create a fasta file with these protein sequences.

First we are going to translate our sequence from our table. To translate we can re-use the package seqinr and see if we can find some useful function with it.

We are going to create again a for loop, to take sequence by sequence, translate it and save it into a fasta file

```{r}

prophage_sequence_table <- read_excel("~/pCloud Drive/thesis/intership_bac_3/prophage_sequence_table.xlsx")

prophage_sequence_table$aa_sequence <- ""

# this loop take sequence by sequence, translate it and save it into fasta file

for(i in seq(from=1, to=nrow(prophage_sequence_table), by=1)){
  
  # we add a print to see the progression
  
  print(paste("Sequence in translation nb : ",i,sep=""))
  
  # We extract one sequence
  
  prophage_sequence_table[i,2] -> one_sequence
  
  # we put it in minuscule (necessary for translate function)
  
  str_to_lower(one_sequence) -> seq_to_trslt

  # we translate it into aa
  
  s2c(seq_to_trslt) -> seq_to_trslt
  translate(seq = seq_to_trslt) -> seq_to_trslt
  
  # we concat all aa into a sequence

  paste(seq_to_trslt,collapse = "") ->  prophage_sequence_table[i,10]
  
  # we save it into a fasta file
  # we give simply the sequence first, then the name of the sequence, the path of the file, and open = a to say to avoid create a new file at each time
  write.fasta(prophage_sequence_table[i,10],prophage_sequence_table[i,1],"~/pCloud Drive/thesis/intership_bac_3/prophage_protein_seq.fasta" , open = "a", nbchar = 60, as.string = FALSE)



} 

# I save again my excel table

write_xlsx(prophage_sequence_table,"/Users/remydugauquier/pCloud\ Drive/thesis/intership_bac_3/prophage_sequence_table.xlsx")

```


# Functionnal prediction analyze of results

With HMMsearch and the PFAM-A database we annotated the function of each genes. 
So we are going to add these function into our table, to reveal the unknown genes.

```{r}
prophage_sequence_table <- read_excel("~/pCloud Drive/thesis/intership_bac_3/prophage_sequence_table.xlsx")


prophage_pfam_result_TAB <- read.table("~/pCloud Drive/thesis/intership_bac_3/prophage_pfam_result_TAB.txt", quote="\"")

# we select column and rename correctly the columns name

prophage_pfam_result_TAB[,c(1,3,4,5,6,7)] -> prophage_pfam_result_TAB

colnames(prophage_pfam_result_TAB) = c("target_name","query_name","query_accession","Evalue_full_seq" ,"score_full_seq","bias_full_seq") 

```

We can't use directly all prediction because some can be very bad prediction.

We need to use two values to filter the table (according to Hmmsearch documentation) :

  - evalue
  - score
  
A good prediction need to have :
  - an evalue lower than 1e-3
  - a bias score lower than the half of the score

So now we subset our table with these parameters

```{r}


```

How many different genes we have with a function?
What the percentage of genes from our contigs with a function?

```{r}


```

It seems that some genes have several identification, so we keep the best by using evalue and score.

```{r}
prophage_pfam_result_TAB_filter %>%
  group_by(target_name) %>%
  top_n(-1,Evalue_full_seq )-> prophage_pfam_result_TAB_filter

# we check if we have got only one identification by genes

length(unique(prophage_pfam_result_TAB_filter$target_name))

```

Now we can merge our table to add these identification

```{r}
# we change the name of the column to have a common column between the two table

names(prophage_pfam_result_TAB_filter)[1] <- "sequence_name"

merge(prophage_sequence_table, prophage_pfam_result_TAB_filter, by="sequence_name",all.x=TRUE) -> prophage_sequence_table

# and we save the table

write_xlsx(prophage_sequence_table,"/Users/remydugauquier/pCloud\ Drive/thesis/intership_bac_3/prophage_sequence_table.xlsx")

```


# Contig figure


```{r}
prophage_sequence_table <- read_excel("~/pCloud Drive/thesis/intership_bac_3/prophage_sequence_table.xlsx")


```

Before the creation of figure, we can check how many different contigs we have.


```{r}
length(unique(prophage_sequence_table$contig_id))
```


We are going to compute a for loop where we assign a color by genes (green if functionnal prediction and red if unknown genes). This loop take contig by contig and compute the figure with Gviz package


```{r}
# first I create a vector with all the contig

unique(prophage_sequence_table$contig_id) -> all_contig

for(i in seq(from=1, to=length(all_contig),by=1)){
  
  # I filter the first contig
  print(i)
  subset(prophage_sequence_table,prophage_sequence_table$contig_id %in% all_contig[i])-> local_contig

  # I change color according to functionnal identification
  local_contig$color <- "#85C17E"
  local_contig$color[is.na(local_contig$query_accession)] <- "#B82010"
  
  # I add an id for the figure
  
  local_contig$gene_order <- rownames(local_contig)
  
  # cwith GRanges we create the back of figure
  ref <- GRanges('candidate_id', IRanges(min(local_contig$start), max(local_contig$stop)))
  ref_track <- GenomeAxisTrack(ref, lwd=4, fontsize=20)
  
  # and then the function to compute arrow according to strand, length, color
  
  local_contig_g <- with(local_contig, GRanges(contig_id, IRanges(start, stop), strand,id=gene_order, fill=color))
  options(ucscChromosomeNames=FALSE)
  local_contig_track <- AnnotationTrack(local_contig_g,
                                        name = unique(local_contig$contig_id)
                                        , width = 25, showFeatureId = T
                                        , min.height=15
                                        , stacking="dense"
                                        , showFeatureId=TRUE
                                        ,id = local_contig_g$id
                                        , fill=local_contig_g$fill)
  

  
  png(file=paste("~/pCloud Drive/thesis/intership_bac_3/figure/",unique(local_contig$contig_id),"_figure.png", sep=""),width=1360, height=220)
  plotTracks(c(ref_track, local_contig_track),title.width = 0.5) -> plot_fig
  dev.off()

}

```





